Miniproject - Algebraic Geometry

### 2021-12-10 Algebraic geometry/submission.tex
Added statement on Gr(2, 6) to alg geo

    By Lemma~\ref{prop:leading_term_ideal_grassmanian} we know
    \begin{equation*}
        J = \mathrm{lt}(\I(\Gr(2, n))) = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R := k[x_{i, j} \ | \ i < j]
    \end{equation*}
    Let $N = {n \choose 2}$ be the number of variables in $R$, i.e. $\{x_1, ..., x_N\} = \{ x_{i, j} \ | \ i < j \}$. Now we find for sufficiently large $m$ that
    \begin{align*}
        &|\{ x^\alpha \ \text{monomial in $R$} \ | \ \deg(x^\alpha) = m, \ \forall i < j < u < v: x_{iv} x_{ju} \ \nmid \ x^\alpha \}| \\
        =\ &|\{ x^\alpha \ \text{monomial in $R$} \ | \ \deg(x^\alpha) = m, \ \forall i < j < u < v: x_{iv} x_{ju} \ \nmid \ \mathrm{sqfr}(x^\alpha) \}| \\
        =\ &\Bigl| \bigcup_{\substack{\alpha \in \{0, 1\}^N\\x_{iv} x_ju \ \nmid \ x^\alpha}} \{ x^\alpha x^\beta \ | \ \deg(x^\beta) = m - \deg(x^\alpha), \ \forall i: \alpha_i = 0 \Rightarrow \beta_i = 0 \} \Bigr| \\
        =& \sum_{\substack{\alpha \in \{0, 1\}^N\\x_{iv} x_ju \ \nmid \ x^\alpha}} \left(\!\!{\deg(x^\alpha) \choose m - \deg(x^\alpha)}\!\!\right) = \sum_{l = 0}^N \sum_{\substack{\alpha \in \{0, 1\}^N\\\sum_j \alpha_j = l\\x_{i, v} x_{j, u} \ \nmid \ x^\alpha}} \left(\!\!{ l \choose m - l }\!\!\right) \\
        =& \sum_{l = 0}^N s_n(l) \left(\!\!{ l \choose m - l }\!\!\right) = \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{align*}
    where $\mathrm{sqrf}(f)$ denotes the square-free part of $f$.
    This holds, as by definition of $s_n$ and $E_n$ we find
    \begin{equation*}
        s_n(l) = |\bigl\{ I \subseteq \{ x_{i, j} \} \ \bigm| \ \forall i < j < u < v: x_{i, v} x_{j, u} \ \nmid \ xy \ \text{for all} \ x, y \in I \bigr\}|
    \end{equation*}
    Now Macaulay's basis theorem~\ref{prop:macaulay_basis_theorem} yields that for sufficiently large $m$ have
    \begin{align*}
        \dim_k (R/\I(\Gr(2, n))) =& |\bigl\{ x^\alpha \ \text{monomial in $R$} \ \bigm| \ \deg(x^\alpha) = m, \ x^\alpha \notin \mathrm{lt}(\I(\Gr(2, n))) \bigr\}| \\
        =& \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{align*}
    Hence we find for the Hilbert polynomial that
    \begin{equation*}
        p_{\Gr(2, n)} = \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{equation*}
    and in particular, it has the leading term $s_n(d) {m - 1 \choose d - 1}$.
    The claim follows by the characterization of degree and dimension using the Hilbert polynomial that we did in the lecture.
\end{proof}
Now we want to study how the independent sets in $G_n$ look like.
\begin{example}
    \label{ex:Gn}
    The graph $G_4$ is the following
    \\\includegraphics[width = \textwidth]{graph4.pdf}
    The graph $G_5$ is the following
    \\\includegraphics[width = \textwidth]{graph5.pdf}
    The graph $G_6$ is the following
    \\\includegraphics[width = \textwidth]{graph6.pdf}
\end{example}
\begin{lemma}
    The largest independent set in $G_n$ is of size $2n - 4$.
\end{lemma}
\begin{proof}
    The idea is to write the graph in layers (or as a kind of ``pyramid'') as displayed in Example~\ref{ex:Gn}.
    Then we can ``push'' an independent set onto the bottom layer, by repeatedly taking vertices in the set that are maximally high up and on the outside, and replacing them by their ``child''. 
    This shows the claim, as the bottom layer (plus the 4 unconnected vertices) has size $2n - 4$.
\end{proof}
\printbibliography
\end{document}

>>>>

    By Lemma~\ref{prop:leading_term_ideal_grassmanian} we know
    \begin{equation*}
        J = \mathrm{lt}(\I(\Gr(2, n))) = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R := k[x_{i, j} \ | \ i < j]
    \end{equation*}
    Let $N = {n \choose 2}$ be the number of variables in $R$, i.e. $\{x_1, ..., x_N\} = \{ x_{i, j} \ | \ i < j \}$. Now we find for sufficiently large $m$ that
    \begin{align*}
        &|\{ x^\alpha \ \text{monomial in $R$} \ | \ \deg(x^\alpha) = m, \ \forall i < j < u < v: x_{iv} x_{ju} \ \nmid \ x^\alpha \}| \\
        =\ &|\{ x^\alpha \ \text{monomial in $R$} \ | \ \deg(x^\alpha) = m, \ \forall i < j < u < v: x_{iv} x_{ju} \ \nmid \ \mathrm{sqfr}(x^\alpha) \}| \\
        =\ &\Bigl| \bigcup_{\substack{\alpha \in \{0, 1\}^N\\x_{iv} x_ju \ \nmid \ x^\alpha}} \{ x^\alpha x^\beta \ | \ \deg(x^\beta) = m - \deg(x^\alpha), \ \forall i: \alpha_i = 0 \Rightarrow \beta_i = 0 \} \Bigr| \\
        =& \sum_{\substack{\alpha \in \{0, 1\}^N\\x_{iv} x_ju \ \nmid \ x^\alpha}} \left(\!\!{\deg(x^\alpha) \choose m - \deg(x^\alpha)}\!\!\right) = \sum_{l = 0}^N \sum_{\substack{\alpha \in \{0, 1\}^N\\\sum_j \alpha_j = l\\x_{i, v} x_{j, u} \ \nmid \ x^\alpha}} \left(\!\!{ l \choose m - l }\!\!\right) \\
        =& \sum_{l = 0}^N s_n(l) \left(\!\!{ l \choose m - l }\!\!\right) = \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{align*}
    where $\mathrm{sqrf}(f)$ denotes the square-free part of $f$.
    This holds, as by definition of $s_n$ and $E_n$ we find
    \begin{equation*}
        s_n(l) = |\bigl\{ I \subseteq \{ x_{i, j} \} \ \bigm| \ \forall i < j < u < v: x_{i, v} x_{j, u} \ \nmid \ xy \ \text{for all} \ x, y \in I \bigr\}|
    \end{equation*}
    Now Macaulay's basis theorem~\ref{prop:macaulay_basis_theorem} yields that for sufficiently large $m$ have
    \begin{align*}
        \dim_k (R/\I(\Gr(2, n))) =& |\bigl\{ x^\alpha \ \text{monomial in $R$} \ \bigm| \ \deg(x^\alpha) = m, \ x^\alpha \notin \mathrm{lt}(\I(\Gr(2, n))) \bigr\}| \\
        =& \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{align*}
    Hence we find for the Hilbert polynomial that
    \begin{equation*}
        p_{\Gr(2, n)} = \sum_{l = 0}^N s_n(l) {m - 1 \choose l - 1}
    \end{equation*}
    and in particular, it has the leading term $s_n(d) {m - 1 \choose d - 1}$.
    The claim follows by the characterization of degree and dimension using the Hilbert polynomial that we did in the lecture.
\end{proof}
Now we want to study how the independent sets in $G_n$ look like.
\begin{example}
    \label{ex:Gn}
    The graph $G_4$ is the following
    \\\includegraphics[width = \textwidth]{graph4.pdf}
    The graph $G_5$ is the following
    \\\includegraphics[width = \textwidth]{graph5.pdf}
    The graph $G_6$ is the following
    \\\includegraphics[width = \textwidth]{graph6.pdf}
\end{example}
\begin{lemma}
    The largest independent set in $G_n$ is of size $2n - 3$.
\end{lemma}
\begin{proof}
    The idea is to write the graph in layers (or as a kind of ``pyramid'') as displayed in Example~\ref{ex:Gn}.
    Then we can ``push'' an independent set onto the bottom layer, by repeatedly taking vertices in the set that are maximally high up and on the outside, and replacing them by their ``child''. 
    This shows the claim, as the bottom layer (plus the 4 unconnected vertices) has size $2n - 3$.
\end{proof}

\begin{corollary}
    Have $\dim(\Gr(2, n)) = 2n - 4$.
\end{corollary}
\begin{proof}
    Follows directly from the previous two statements.
\end{proof}

\begin{corollary}
    Have $\dim(\Gr(2, 6)) = 8$ and $\deg(\Gr(2, 6)) = 14$.
\end{corollary}
\begin{proof}
    The dimension follows directly from the previous general case.
    For the degree, just count the maximal independent sets in $G_6$, as displayed in Example~\ref{ex:Gn}.
    This is slightly tricky, but it is not too hard to see that there are exactly 15 of them.
\end{proof}
Of course, the above statement on $\Gr(2, 6)$ can also easily be found using Computer Algebra.
For example, the following Sage script shows that
\begin{equation*}
    p_{\Gr(2, 6)} = \frac 1 {2880} t^8 + \frac 1 {120} t^7 + \frac {41} {480} t^6 + \frac {39} {80} t^5 + \frac {541} {320} t^4 + \frac {291} {80} t^3 + \frac {3401} {720} t^2 + \frac {101} {30} t + 1
\end{equation*}
from which we easily deduce that
\begin{equation*}
    \dim(\Gr(2, 6)) = 8 \quad \text{and} \quad \deg(\Gr(2, 6)) = \frac 1 {2880} \cdot 8! = 14
\end{equation*}
\lstinputlisting[language = python]{./gr26.sage}
\printbibliography
\end{document}
